---
title: "Probabilistic data linkage tutorial"
author: "Cecina Babich Morrow"
date: "2026-05-27"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Probabilistic data linkage

This tutorial demonstrates how to perform probabilistic data linkage in R using `fastLink`.


## Install and load necessary packages

First, ensure that you have `fastLink` and the `tidyverse` suite installed and loaded:

```{r}
# Install fastLink if you haven't already
if (!requireNamespace("fastLink", quietly = TRUE)) {
  install.packages("fastLink")
}
# Load the fastLink package
library(fastLink)

# Install tidyverse if you haven't already
if (!requireNamespace("tidyverse", quietly = TRUE)) {
  install.packages("tidyverse")
}
# Load the tidyverse package
library(tidyverse)
```

## Toy dataset

Let's start by looking at the probabilities of match on the very small toy dataset from the presentation.

```{r}
# Load the dataset
example_data_1 <- read_csv("data/example-data-1.csv")

# View the dataset
print(example_data_1)
```

We can run `fastLink` on this dataset to see the probabilities of match for each pair of records. Don't worry too much about what all the arguments to the `fastLink` function mean at this stage -- we will go through them in more detail later.

```{r}
# Run fastLink on the dataset
example_fl_1 <- fastLink(
  dfA = example_data_1,
  dfB = example_data_1,
  varnames = colnames(example_data_1),
  stringdist.match = colnames(example_data_1)
)

# Print the probabilities that each pair of records is a match
print(example_fl_1$matches)
print(example_fl_1$posterior)
```

## `samplematch` dataset

Now we can use a larger dataset to see how to implement probabilistic data linkage in practice. We will use the `samplematch` dataset to explore. Add some code to the chunk below to examine the two preloaded dataframes, which are named `dfA` and `dfB`:

```{r}
# Load the built-in data
data(samplematch)

# YOU TRY:
# View the first few rows of dfA and dfB

# YOU TRY:
# Print the column names of each dataframe

```

### Basic `fastLink` arguments

The `fastLink` function has several arguments that allow you to customize the data linkage process. Here are some of the key arguments:

- `dfA` and `dfB`: The two dataframes you want to link.
- `varnames`: A vector of column names that you want to use for matching. These should be the same in both dataframes.
- `stringdist.match`: A vector of column names for which you want to use string distance matching. This is useful for columns that may have typos or variations in spelling.

Note that for any variables included in `varnames`, but not in `stringdist.match`, `fastLink` will consider two records to match on that variable if the values are exactly the same. For variables included in `stringdist.match`, `fastLink` will calculate a string similarity between the values and consider them a match if the distance is above a certain threshold.

Let's start by running a preliminary linkage using `fastLink` and the set of arguments listed above:

```{r}
# YOU TRY:
matches1 <- fastLink(
  
)
```

We can summarise the results of this preliminary linkage:

```{r}
summary(matches1)
```

Let's investigate the different components of the summary output:

+ The columns: these represent different cutoffs for the posterior probabilities of match.
  + For example, the column `85%` shows the number of matches that have a posterior probability of match greater than or equal to 0.85. 
  + The `Exact` column shows the number of matches that have a posterior probability of match equal to 1, which means that they are considered an exact match.
+ `Match Count`: the number of pairs of records that are considered matches at each cutoff.
+ `Match Rate`: the proportion of pairs of records that are considered matches at each cutoff, out of all possible pairs of records.
+ `FDR`: the false discovery rate, which is the expected proportion of false matches among the pairs of records that are considered matches at each cutoff.
+ `FNR`: the false negative rate, which is the expected proportion of true matches that are not identified as matches at each cutoff.

Next, extract the matches from the `matches1` object using the `getMatches()` function and view the first few rows:

```{r}
# YOU TRY:
matches1_extracted <- getMatches(

)
# View the first few rows:

```

We can also view the distribution of posterior probabilities, i.e. the probabilities that each pair of records is a match:

```{r}
# YOU TRY:

```

What can you tell from this histogram? Is it particularly informative?

### String distance arguments

Beyond the basic arguments to the `fastLink` function that we used above, we have even more options to customize the data linkage process. Currently, when we use `stringdist.match`, we consider two records to match on a given variable if the string similarity is greater than or equal to 0.94. There are several arguments that can adjust the way we perform string matching:

+ `partial.match`: A vector of column names for which you want to use partial string matching. This is useful for columns that may have variations in formatting (e.g., "St" vs "Street").
+ `cut.a`: A numeric value between 0 and 1 that sets the threshold for considering two records a full match based on string similarity. The default is 0.94, but you can adjust this based on the level of similarity you want to require for a match.
+ `cut.p`: A numeric value between 0 and 1 that sets the threshold for considering two records a partial match based on string similarity. The default is 0.88, but you can adjust this based on the level of similarity you want to require for a partial match.

Try re-running the `fastLink` function using `partial.match`, and adjusting the `cut.a` and `cut.p` thresholds to see how it affects the results:

```{r}
# YOU TRY:
matches2 <- fastLink(
  
)

# Summarise results:


# Extract matches:


# View distribution of posterior probabilities:

```

### Numeric matching

In addition to matching on string variables, we can also match on numeric variables. For example, we might want to match on `birthyear` using a numeric distance rather than a string distance, if we think it is possible that a birth year might be entered one off from the actual value, for example. The arguments pertaining to numeric matching are:

+ `numeric.match`: A vector of column names for which you want to use numeric distance matching. This is useful for columns that contain numeric values where small differences may still indicate a match (e.g., birth year).
+ `cut.a.num`: A numeric value that sets the threshold for considering two records a full match based on numeric distance. The default is 1, which means that two records will be considered a full match if the absolute difference between the values is less than or equal to 1.
+ `cut.p.num`: A numeric value that sets the threshold for considering two records a partial match based on numeric distance. The default is 2.5, which means that two records will be considered a partial match if the absolute difference between the values is less than or equal to 2.5.

Try re-running the `fastLink` function using `numeric.match`, and adjusting the `cut.a.num` and `cut.p.num` thresholds to see how it affects the results:

```{r}
# YOU TRY:
matches3 <- fastLink(
  
)

# Summarise results:


# Extract matches:


# View distribution of posterior probabilities:

```

### Reweighting names

If two people have the same first name, the amount of evidence we have that they are a match depends on how common that first name is in the population. For example, my first name is "Cecina". Since I have never met another Cecina, if I see two records with the first name "Cecina", that is strong evidence that those two records are a match. On the other hand, "Chris" is a relatively common name, so two records with the first name "Chris" are less likely to be a match than two records with the first name "Cecina". The following arguments allow us to reweight the evidence of a match based on the frequency of the first name in the population:

+ `reweight.names`: A logical value indicating whether to reweight the evidence of a match based on the frequency of the first name in the population. The default is `FALSE`, but you can set it to `TRUE` to enable this feature.
+ `firstname.field`: A character string specifying the name of the column that contains the first names in your data. This is required if you set `reweight.names` to `TRUE`.

Try re-running the `fastLink` function using `reweight.names` to see how it affects the results:

```{r}
# YOU TRY:
matches4 <- fastLink(
  
)

# Summarise results:


# Extract matches:


# View distribution of posterior probabilities:

```

